Modeling Lake Trophic State: A Data Mining Approach

### 2014-09-12 inst/doc/manuscript.Rmd
cleaned up vignettes, got table and figure labels and captions working, rebuilt example in inst/doc

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author:
- affiliation: aed
  email: hollister.jeff(at)epa.gov
  footnote: Corresponding author
  name: Jeffrey W. Hollister
- affiliation: aed
  name: W. Bryan Milstead
- affiliation: aed
  name: Betty J. Kreakie
address:
- address: "US Environmental Protection Agency Office of Research and Development National
    Health and Environmental Effects Research Laboratory Atlantic Ecology Division 27
    Tarzwell Drive  Narragansett, RI, 02879, USA  "
  code: aed
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: elsarticle.latex
  html_document: null
  word_document: null
fontsize: 11pt
csl: plos.csl
documentclass: elsarticle
layout: review
linenumbers: yes
bibliography: LakeTrophicModelling.bib
abstract:  Productivity of lentic ecosystems has been well studied and it is 
  widely accepted that as nutrient inputs increase,  productivity increases and
  lakes transition from low trophic state (e.g. oligotrophic) to higher trophic 
  states  (e.g. eutrophic). These broad trophic state classifications are good 
  predictors of ecosystem health and ecosystem services/disservices (e.g. recreation,
  aesthetics, fisheries, and harmful algal blooms). While the relationship between 
  nutrients and trophic state provides reliable predictions, it requires *in situ* 
  water quality data in order to paramterize the model.  This limits the application 
  of these models to lakes with existing and, more importanly, available water quality
  data.  To expand our ability to predict in lakes without water quality data, we take
  advantage of the availability of a large national lakes water quality database, 
  land use/land cover data, lake morphometry data, other universally available data, 
  and modern data mining approaches to build and assess models of lake tropic state 
  that may be more universally applied.  We use random forests and random forest 
  variable selection to identify variables to be used for predicting trophic state and 
  we compare the performance of two models of trophic state (as determined by 
  chlorophyll a concentration).   The first model estimates trophic state with *in situ* 
  as well as universally available data and the second model uses universally available 
  data only. For each of these models we used three separate trophic state categories, 
  for a total of six models. Overall accuracy for the *in situ* and universal data 
  models ranged from xx% to xx% and xx, xx, and xx described the most variation in 
  trophic state.  For the universal data only models, Overall accuraccy ranged from xx% 
  to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it 
  is believed that the presence and abundance of cyanobacteria is strongly associated 
  with trophic state.  To test this we examine the association between estimates of 
  cyanobacteria biovolume and the measured and predicted trophic state.  Expanding 
  these preliminary results to include cyanobacteria taxa indicates that cyanobacteria 
  are significantly more likely to be found in highly eutrophic lakes. These results 
  suggest that predictive models of lake trophic state may be improved with additional 
  information on the landscape surrounding lakes and that those models provide 
  additional information on the presence of potentially harmful cyanobacteria taxa. 
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}

-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
  if(!before)
    paste('\n\n:', options$tab.cap, sep='') 
})
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == F)  
    x
  else
    default_output_hook(x,options)
})
```

@@

The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 locations across the conterminous United States (Map 1), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake water quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].  Adding to the monitoring data collected via the NLA, we use the 2006 NLCD data to examine the possible landscape-level drivers of trophic status in lakes.  The NLCD is a nationally collected land use land cover dataset that also provides estimates of impervious surface.  We collected total land use land cover and total percent impervious surface within a 3 kilometer buffer surrounding the lake to examine larger landscape-level effect [@homer2004development;@xian2009updating]. We also used various measures of lake morphometry (i.e. depth, volume, fetch, etc.) as they are important in understanding lake productivity, yet many of these data are difficult to obtain for large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@NLMDInPrep]. Lastly, to explore associations between trophic state and cyanobacteria, we used estimates of cyanobacterial biovoulme caluclated by Beaulieu *et al.* [-@beaulieu2013nutrients].  Cyanobacteria biovolumes are a truer measure of cyanobacteria dominance than abundance as there is great variability in the size within and between species.  We have consolidated the taxa level estimates from Beaulieu *et al.* [-@beaulieu2013nutrients] and  summed that information on a per-lake basis.

@@

```{r results='asis', echo=FALSE, fig.cap="Map of the distribution of National Lakes Assesment Sampling locations"}
state<-map_data('state')
lakes_alb<-data.frame(hkm2014Data[["AlbersX"]],hkm2014Data[["AlbersY"]])
p4s<-"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
ll<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
lakes_alb_sp<-SpatialPoints(coordinates(lakes_alb),proj4string=CRS(p4s))
lakes_dd<-spTransform(lakes_alb_sp,CRS=CRS(ll))
lakes_dd<-data.frame(coordinates(lakes_dd))
mycolor<-wes.palette(5,"Zissou2")
mycolor<-c(mycolor[1],mycolor[2],mycolor[4])
names(lakes_dd)<-c("long","lat",c())
nlaMap(state,lakes_dd,mycolor)
```

**Predicting Trophic State with Random Forests**

@@

**Model Details**

Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table 2) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

@@

```{r trophicStateTable, results='asis', echo=FALSE, tab.cap="Chlorphyll a based trophic state cut-offs"}
ts_4<-c("oligo","meso","eu","hyper")
ts_3<-c("oligo","meso/eu","meso/eu","hyper")
ts_2<-c("oligo/meso","oligo/meso","eu/hyper","eu/hyper")
co<-c("<= 0.2",">2-7",">7-30",">30")
xdf<-data.frame(ts_4,ts_3,ts_2,co)
names(xdf)<-c("Trophic State (4)","Trophic State (3)","Trophic State (2)","Cut-off")
kable(xdf)
```

##Results

```{r figSetup,echo=FALSE,eval=T,cache=TRUE}
#create list of all selected variables and assign a color
vars<-ls(pattern="_vars")
all_vars<-vector()
for(i in vars){
  all_vars<-c(all_vars,get(i))
}
all_vars<-unique(all_vars)

@@

###Model 1: 4 Trophic States ~ All Variables

@@

Total accuracy for Model 1 is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,3)`.

```{r VarSel_Model1,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 1"}
kable(sumTable(all_ts4[1:100]),row.names=F)
```

```{r Confusion_Model1,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 1"}
kable(formatC(round(all_ts4_rf$confusion,2)),row.names=F)
```

```{r Importance_Model1,results="asis",echo=FALSE, fig.cap="Importance plot for Model 1"}
all_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts4_vars]
importancePlot(all_ts4_rf,'gini',size=10,aes(colour=all_ts4_color))
```

###Model 2: 3 Trophic States ~ All Variables

@@

###Model 3: 2 Trophic States ~ All Variables

@@

###Model 4: 4 Trophic States ~ GIS Only Variables

@@

```{r Importance_Model4,results="asis",echo=FALSE, tab.cap="Importance plot for Model 3"}
gis_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts4_vars]
importancePlot(gis_ts4_rf,'gini',size=10,aes(colour=gis_ts4_color))
```

###Model 5: 3 Trophic States ~ GIS Only Variables

@@

###Model 6: 2 Trophic States ~ GIS Only Variables

@@

###Associating Trophic State and Cyanobacteria

```{r ts_4_biov, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 4 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_4,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:9],
        y='Percent',x='Log10(Cyanobacteria Biovolume)',
        title=expression(paste('CDF for Chlorophyll ', 
                                italic("a"),' Trophic States (4 Categories)')),
        color="Trophic State\nCategories")
```

```{r ts_3_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 3 trophic state classes" }
plotCdf(hkm2014Data$TS_CHLA_3,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:8],
       y='Percent',x='Log10(Cyanobacteria Biovolume)', 
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (3 Categories)')),
       color="Trophic State\nCategories")
```

```{r ts_2_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 2 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_2,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:7],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (2 Categories)')),
       color="Trophic State\nCategories")
```

```{r scatterplot,echo=FALSE, fig.width=7, fig.cap="Cholorphyll *a* and cyanobacteria biovolume scatterplot"}
scp_df<-data.frame(chla=hkm2014Data[["CHLA"]],biovp1=hkm2014Data[["sumBioV"]]+1)
scatterPlot(scp_df,xvar="chla",yvar="biovp1",zissou[8],zissou[7],zissou[6],
             title=expression(paste("Chlorophyll ", 
                                    italic("a")," and Cyanobacteria Relationship")),
             x=expression(paste('Log10(Chl ', italic("a"),')')),
             y="Log10(Cyanobaterial Biovolumes + 1)")
```

>>>>

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author:
- affiliation: aed
  email: hollister.jeff(at)epa.gov
  footnote: Corresponding author
  name: Jeffrey W. Hollister
- affiliation: aed
  name: W. Bryan Milstead
- affiliation: aed
  name: Betty J. Kreakie
address:
- address: "US Environmental Protection Agency Office of Research and Development National
    Health and Environmental Effects Research Laboratory Atlantic Ecology Division 27
    Tarzwell Drive  Narragansett, RI, 02879, USA  "
  code: aed
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/elsarticle.latex
  html_document: null
  word_document: null
fontsize: 11pt
csl: components/ecology.csl
documentclass: elsarticle
layout: review
linenumbers: no
bibliography: components/LakeTrophicModelling.bib
abstract:  Productivity of lentic ecosystems has been well studied and it is 
  widely accepted that as nutrient inputs increase,  productivity increases and
  lakes transition from low trophic state (e.g. oligotrophic) to higher trophic 
  states  (e.g. eutrophic). These broad trophic state classifications are good 
  predictors of ecosystem health and ecosystem services/disservices (e.g. recreation,
  aesthetics, fisheries, and harmful algal blooms). While the relationship between 
  nutrients and trophic state provides reliable predictions, it requires *in situ* 
  water quality data in order to paramterize the model.  This limits the application 
  of these models to lakes with existing and, more importanly, available water quality
  data.  To expand our ability to predict in lakes without water quality data, we take
  advantage of the availability of a large national lakes water quality database, 
  land use/land cover data, lake morphometry data, other universally available data, 
  and modern data mining approaches to build and assess models of lake tropic state 
  that may be more universally applied.  We use random forests and random forest 
  variable selection to identify variables to be used for predicting trophic state and 
  we compare the performance of two models of trophic state (as determined by 
  chlorophyll a concentration).   The first model estimates trophic state with *in situ* 
  as well as universally available data and the second model uses universally available 
  data only. For each of these models we used three separate trophic state categories, 
  for a total of six models. Overall accuracy for the *in situ* and universal data 
  models ranged from xx% to xx% and xx, xx, and xx described the most variation in 
  trophic state.  For the universal data only models, Overall accuraccy ranged from xx% 
  to xx% and xx, xx, and xx described the most variation in trophic state. Lastly, it 
  is believed that the presence and abundance of cyanobacteria is strongly associated 
  with trophic state.  To test this we examine the association between estimates of 
  cyanobacteria biovolume and the measured and predicted trophic state.  Expanding 
  these preliminary results to include cyanobacteria taxa indicates that cyanobacteria 
  are significantly more likely to be found in highly eutrophic lakes. These results 
  suggest that predictive models of lake trophic state may be improved with additional 
  information on the landscape surrounding lakes and that those models provide 
  additional information on the presence of potentially harmful cyanobacteria taxa. 
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Modeling Lake Trophic State: A Data Mining Approach}
-->

@@

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})

```

@@

The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 locations across the conterminous United States (Figure \ref{fig:nlaMap}), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake water quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].  Adding to the monitoring data collected via the NLA, we use the 2006 NLCD data to examine the possible landscape-level drivers of trophic status in lakes.  The NLCD is a nationally collected land use land cover dataset that also provides estimates of impervious surface.  We collected total land use land cover and total percent impervious surface within a 3 kilometer buffer surrounding the lake to examine larger landscape-level effect [@homer2004development;@xian2009updating]. We also used various measures of lake morphometry (i.e. depth, volume, fetch, etc.) as they are important in understanding lake productivity, yet many of these data are difficult to obtain for large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@NLMDInPrep]. Lastly, to explore associations between trophic state and cyanobacteria, we used estimates of cyanobacterial biovoulme caluclated by Beaulieu *et al.* [-@beaulieu2013nutrients].  Cyanobacteria biovolumes are a truer measure of cyanobacteria dominance than abundance as there is great variability in the size within and between species.  We have consolidated the taxa level estimates from Beaulieu *et al.* [-@beaulieu2013nutrients] and  summed that information on a per-lake basis.

@@

```{r nlaMap, echo=FALSE, fig.cap="Map of the distribution of National Lakes Assesment Sampling locations\\label{fig:nlaMap}"}
state<-map_data('state')
lakes_alb<-data.frame(hkm2014Data[["AlbersX"]],hkm2014Data[["AlbersY"]])
p4s<-"+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
ll<-"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
lakes_alb_sp<-SpatialPoints(coordinates(lakes_alb),proj4string=CRS(p4s))
lakes_dd<-spTransform(lakes_alb_sp,CRS=CRS(ll))
lakes_dd<-data.frame(coordinates(lakes_dd))
mycolor<-wes.palette(5,"Zissou2")
mycolor<-c(mycolor[1],mycolor[2],mycolor[4])
names(lakes_dd)<-c("long","lat",c())
nlaMap(state,lakes_dd,mycolor)
```

##Predicting Trophic State with Random Forests

@@

##Model Details

Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table \ref{tab:trophicStateTable}) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

@@

```{r trophicStateTable, results='asis', echo=FALSE, tab.cap="Chlorophyll a based trophic state cut-offs\\label{tab:trophicStateTable}"}
ts_4<-c("oligo","meso","eu","hyper")
ts_3<-c("oligo","meso/eu","meso/eu","hyper")
ts_2<-c("oligo/meso","oligo/meso","eu/hyper","eu/hyper")
co<-c("<= 0.2",">2-7",">7-30",">30")
xdf<-data.frame(ts_4,ts_3,ts_2,co)
names(xdf)<-c("Trophic State (4)","Trophic State (3)","Trophic State (2)","Cut-off")
kable(xdf)
```

#Results

```{r figSetup,echo=FALSE,eval=TRUE,cache=TRUE}
#create list of all selected variables and assign a color
vars<-ls(pattern="_vars")
all_vars<-vector()
for(i in vars){
  all_vars<-c(all_vars,get(i))
}
all_vars<-unique(all_vars)

@@

##Model 1: 4 Trophic States ~ All Variables

@@

The selected variables that made up Model 1 were Potassium, Nitrogen:Phosphorus, Total Nitrogen, Total Phosphorus, Total Organic Carbon, Turbidity, Ecoregion, Organic Ions, Dissolved Organic Carbon, and Maximum Lake Depth (Table \ref{tab:VarSel_Model1}). Total accuracy for Model 1 is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$diag,3)`% and the Cohen's Kappa is `r round(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,3)` (Table \ref{tab:Confusion_Model1}).

```{r VarSel_Model1,results="asis",echo=FALSE, tab.cap="Variable selection results for Model 1\\label{tab:VarSel_Model1}"}
kable(sumTable(all_ts4[1:100]),row.names=F)
```

```{r Confusion_Model1,results="asis",echo=FALSE, tab.cap="Random Forest confusion matrix for Model 1\\label{tab:Confusion_Model1}"}
kable(formatC(round(all_ts4_rf$confusion,2)),row.names=F)
```

Lastly, tubidity, total phosphorus, total nitrogen, and total organic carbon were the most important predictors of the 4 classes of trophic state (Figure \ref{fig:Importance_Model1}).

```{r Importance_Model1,results="asis",echo=FALSE, fig.cap="Importance plot for Model 1\\label{fig:Importance_Model1}"}
all_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%all_ts4_vars]
importancePlot(all_ts4_rf,'gini',size=10,aes(colour=all_ts4_color))
```

##Model 2: 3 Trophic States ~ All Variables

@@

##Model 3: 2 Trophic States ~ All Variables

@@

##Model 4: 4 Trophic States ~ GIS Only Variables

@@

```{r Importance_Model4,results="asis",echo=FALSE, tab.cap="Importance plot for Model 4"}
gis_ts4_color<-col_lu[["hexcode"]][col_lu[["variables"]]%in%gis_ts4_vars]
importancePlot(gis_ts4_rf,'gini',size=10,aes(colour=gis_ts4_color))
```

##Model 5: 3 Trophic States ~ GIS Only Variables

@@

##Model 6: 2 Trophic States ~ GIS Only Variables

@@

##Associating Trophic State and Cyanobacteria

```{r ts_4_biov, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 4 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_4,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:9],
        y='Percent',x='Log10(Cyanobacteria Biovolume)',
        title=expression(paste('CDF for Chlorophyll ', 
                                italic("a"),' Trophic States (4 Categories)')),
        color="Trophic State\nCategories")
```
\newpage
```{r ts_3_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 3 trophic state classes" }
plotCdf(hkm2014Data$TS_CHLA_3,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:8],
       y='Percent',x='Log10(Cyanobacteria Biovolume)', 
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (3 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r ts_2_biov_cdf, echo=FALSE, fig.width=7, fig.cap="Cumulative distribution function of cyanobacetria biovolume for 2 trophic state classes"}
plotCdf(hkm2014Data$TS_CHLA_2,hkm2014Data$sumBioV+1,cdf_colors=zissou[6:7],
       y='Percent',x='Log10(Cyanobacteria Biovolume)',
       title=expression(paste('CDF for Chlorophyll ', 
                              italic("a"),' Trophic States (2 Categories)')),
       color="Trophic State\nCategories")
```
\newpage
```{r scatterplot,echo=FALSE, fig.width=7, fig.cap="Cholorphyll *a* and cyanobacteria biovolume scatterplot"}
scp_df<-data.frame(chla=hkm2014Data[["CHLA"]],biovp1=hkm2014Data[["sumBioV"]]+1)
scatterPlot(scp_df,xvar="chla",yvar="biovp1",zissou[8],zissou[7],zissou[6],
             title=expression(paste("Chlorophyll ", 
                                    italic("a")," and Cyanobacteria Relationship")),
             x=expression(paste('Log10(Chl ', italic("a"),')')),
             y="Log10(Cyanobaterial Biovolumes + 1)")
```
